---
title: "Elements of Statistical Learning Practice"
author: "Brittney Bailey"
date: "`r format(Sys.time(), '%d %B, %Y')`" 
urlcolor: blue
linkcolor: blue
output:
  pdf_document:
    fig_height: 2
    fig_width: 3.5
    number_sections: yes
    latex_engine: xelatex
classoption: fleqn
# mainfont: Roboto Light
# sansfont: Roboto Light
# monofont: Roboto Mono
#geometry: "top=0.5in,bottom=0.5in"
header-includes:
  - \usepackage{fancyhdr}
#  - \pagenumbering{gobble}
---

```{r setup, include=FALSE}
# load packages
library(tidyverse)
library(kableExtra)

library(glmertree) # partykit libcoin mvtnorm


# set some code chunk defaults
knitr::opts_chunk$set(
  # code defaults
  tidy = FALSE, size = "tiny",
  # indent plain R output
  comment = "\t",
  # center figures
  fig.align = "center", 
  # suppress warnings and messages
  warning = FALSE, message = FALSE) 
# set black & white plot theme as default
theme_set(theme_classic()) 
# slower to display scientific notation; "NA"s knit as blank spaces
options(scipen = 1, knitr.kable.NA = '')
```


# Independent outcomes

## From [`tidymodels`: *Get Started*](https://www.tidymodels.org/start/) 

1. Beginning: The `recipes` package is for pre-processing the data
2. Middle step: The `parsnip` package lets us do all the model fitting and summarizing.
3. Middle step: combine preprocessing and model into workflow using `workflow` package.
4. Pipe the workflow into the fit function with a specified dataset to fit the model.
5. 


### 0. Loading packages and data
The `tidymodels` package relies mostly on `parsnip` package for functions to specify and train models with different "engines."

Before that, we can use the `recipes` package to pre-process the data.


Recipes of the form `recipe(formula, data)` are ordered pre-processing steps applied to data before analysis (formula cannot use in-line functions, data transformations, or minus signs).

1. specify `recipe(formula, data)` where the formula prescribes predictors on the RHS and outcomes on the LHS
    * can add/update roles later as needed using `update_role()`; when you change a role from a predictor or outcome, then the variable will be retained but not included in the model (need to verify!).
    
2. Implement `step_` functions in desired order, making sure to remove any variables from which you derive features, create dummy variables from categorical if needed, make binary outcomes factors if needed, 
    * useful selectors: has_role(),  has_type(), all_numeric(), all_nominal(), all_predictors(), and all_outcomes() plus tidyselect helpers
    * impute:`bag_impute, imp_vars, impute_linear, knnimpute, lowerimpute, meanimpute, medianimpute, modeimpute, rollimpute, unkown`
    * individual transformations: `BoxCox, bs` (B-spline basis), `hyperbolic, inverse, invlogit, log, logit, mutate, ns` (natural spline basis), `poly` (orthogonal polynomial basis), `relu` (smoothed rectified linear transformation), `sqrt, YeoJohnson`
    * discretization: `discretize, cut`
    * dummy variables and encodings: `bin2factor, count, date, dummy, factor2string, holiday, integer, novel, num2factor, ordinalscore, other, regex, relevel, string2factor, unknown,, unorder`
    * interactions: `interact`
    * normalization: `center, normalize, range, scale`
    * multivariate transformations: `classdist, depth, geodist, ica` (ICA signal extraction), `isomap, kpca, kpca_poly, kpca_rbf, mutate_at, nnmf, pca, pls, ratio, spatialsign`
    * filters: `corr, lincomb, nzv, rm, zv`
    * row operations
    * others: `intercept, profile, rename, rename_at, window`
    


### 2. Model fitting, tuning, and summarizing

1. **`_type_(args)`**: Use `parsnip` package to specify *type* of model (`boost_tree, decision_tree(), linear_reg(), logistic_reg(), mars(), mlp(), multinom(), nearest_neighbor(), null_model(), rand_forest(), surv_reg(), svm_poly(), svm_rbf()`)
    * `rand_forest`: `mtry` = number of predictors, `trees` = number of trees, `min_n` = min number of points in a node
    * `logistic_reg`/`linear_reg`: `penalty` = total amount of regularization for glmnet, keras, spark, `mixture` = mixture amounts of different types of regularization 
    * `decision_tree`: `cost_complexity` = the cost/complexity or $C_p$ used by CART models in `rpart`, `tree_depth` = maximum depth of tree in rpart and spark, `min_n` = min number of points in a node
  

2. **`set_mode()`**: Specify the *mode* of the model (e.g., classification, regression, censored regression, risk regression, clustering)
    * `rand_forest`: "unknown", "regression", "classification"
    * `logistic_reg`: "classification"
    * `linear_reg`: "regression"
    * `decision_tree`: "unknown", "regression", "classification"
  
3. **`set_engine()`**:  Specify the *engine* for the model, which is the package or external method used to fit the model (can use `shown_engines("_TYPE_")` to see current set)
    * `rand_forest`: "ranger", "randomForest", "spark" (use `fit_xy()`)
    * `logistic_reg`/`linear_reg`: "glm"/"lm", "glmnet", "stan", "spark", "keras"
    * `decision_tree`: "rpart", "C5.0", "spark"
    
4. If not using workflow, **`fit(y ~ x1 + x2, data = _DATA_)`**: *fit* prescribed model formula

### 3. Combine recipes and model fitting into workflow that can be applied to any dataset

```{r tm-wf-example, eval = FALSE}
flights_wf <- 
  workflow() %>%
  add_model(glm_spec) %>%
  add_recipe(flights_rec)
```



### 4. Fit model to training data and summarize


### 5. Make predictions
### 6. Evaluate model 

We can use roc_curve and roc_auc from the `yardstick` package.

## Tidy models part 3
> The cells data has class labels for 2019 cells — each cell is labeled as either poorly segmented (PS) or well-segmented (WS). Each also has a total of 56 predictors based on automated image analysis measurements. For example, avg_inten_ch_1 is the mean intensity of the data contained in the nucleus, area_ch_1 is the total size of the cell, and so on (some predictors are fairly arcane in nature).


```{r tm-wrangle}
library(tidymodels) # for the rsample package, along with the rest of tidymodels

# Helper packages
library(modeldata)  # for the cells data

# load data
data(cells, package = "modeldata")
```

```{r explore}
cells %>% 
  count(class) %>% 
  mutate(prop = n/sum(n))
```
```{r tm-split-data}
set.seed(123)

cells_split <- initial_split(data = cells %>%
                               select(-case), 
                             prop = 3/4, strata = )

cells_train <- training(cells_split)
cells_test <- testing(cells_split)

# check split
ggplot(cells_train, )
```


```{r tm-recipes}
```




```{r tm-workflow}
# specify model
glm_spec <- logistic_reg() %>%
  set_engine("glm") 

# create workflow that can be applied to any dataset
flights_wf <- 
  workflow() %>%
  add_model(glm_spec) %>%
  add_recipe(flights_rec)
```

```{r tm-fit-summarize}

```



### 5. Make predictions
  
```{r tm-predictions}

```


### 6. Evaluate model

We can use roc_curve and roc_auc from the `yardstick` package.
```{r tm-roc-curve}
```






## From ISLR Lab 8.3: Decision Trees (p. 323)
```{r ISLR1}
# load packages
library(tree)
library(ISLR)
```
```{r tidymodels}
# based on http://www.rebeccabarter.com/blog/2020-03-25_machine_learning/
library(tidymodels)
library(workflows)
library(tune)
library(modeldata)
# from tidymodels
library(readr)
library(broom.mixed)
library(dotwhisker)
```

### Carseats dataset
The *Carseats* dataset has 400 stores of where 11 variables were measured on car seat sales: 3 categorical and 8 quantitative. We want to predict whether the sales are high, meaning greater than 8 thousand at a store. 
```{r include = FALSE}
# load and explore data
data(Carseats)
dim(Carseats)
names(Carseats)
#skimr::skim(Carseats)
```

We start by creating the `highsales` binary variable.
```{r get-data}
# create binary sales variable (make sure it's a factor)
carseats <- mutate(Carseats, 
                   highsales = factor(ifelse(Sales <= 8, "no", "yes")))
```


```{r tidy-preprocessing}
# split data
set.seed(52)
# run splitting function
carseats.split <- rsample::initial_split(carseats, prop = 0.75)

# get training set
carseats.train <- training(carseats.split)

# get test set
carseats.test <- testing(carseats.split)

# get cross-validated version of training set
carseats.cv <- vfold_cv(carseats.train)
```

We can then fit a single classification tree using `tree()` in the `tree` package:
```{r islr-fit-tree}
# fit classification tree
tree.carseats <- tree(highsales~ . - Sales, data = carseats)
summary(tree.carseats)
```

```{r compare-tree-defaults, eval = FALSE, echo = FALSE}
tree.carseats.gini <- tree(highsales~ . - Sales, data = carseats, 
                        split = "gini")
tree.carseats.deviance <- tree(highsales~ . - Sales, data = carseats, 
                            split = "deviance")
summary(tree.carseats.gini)
summary(tree.carseats.deviance)
```

Note the following:  

* `.-Sales` means *regress on everything except the Sales variable*; 
* the default splitting method is based on the *deviance* (alternative: `gini`), $-2\sum_m\sum_kn_{mk}\log(\hat p_{mk})$ where $n_{mk}$ is the number of observations in the $m$th terminal node that belong to class $k$; small deviance is good; and 
* the misclassification error rate provided is the *training* error rate 
* the residual mean deviance is the deviance divided by $n - |T_0|$

```{r tidy-workflow, eval = FALSE}
# specify the model inputs (and any preprocessing)
carseats.recipe <- recipe(highsales ~ . , data = carseats) %>%
  step_rm(Sales)

# prep data
carseats.prepped <- prep(carseats.recipe, carseats) %>%
  juice()

# specify the algorithm
carseats.tree <- 
  decision_tree() %>%
  set_engine("rpart") %>% # cost_complexity (cp)
  set_mode("classification")

# put model and recipes together into workflow
carseats.workflow <-
  workflow() %>%
  add_model(carseats.tree)

# tune parameters if needed (not needed here)

# finalize workflow after tuning parameters (if needed)

# get training model
carseats.tree.fit <- last_fit(carseats.workflow, carseats.split)

```


We can display the tree using `plot()` 
```{r islr-display}
plot(tree.carseats)
```
# Correlated binary outcomes

## Depression data

```{r}
data("DepressionDemo")
```
Simulated dataset of a randomized clinical trial (N = 150) to illustrate fitting of (G)LMM trees.


A data frame containing 150 observations on 6 variables:

`depression` 
numeric. Continuous treatment outcome variable (range: 3-16, M = 9.12, SD = 2.66).

`treatment` 
factor. Binary treatment variable.

`cluster` 
factor. Indicator for cluster with 10 levels.

`age` 
numeric. Continuous partitioning variable (range: 18-69, M = 45, SD = 9.56).

`anxiety` 
numeric. Continuous partitioning variable (range: 3-18, M = 10.26, SD = 3.05).

`duration` 
numeric. Continuous partitioning variable (range: 1-17, M = 6.97, SD = 2.90).

`depression_bin` 
factor. Binarized treatment outcome variable (0 = recovered, 1 = not recovered).

The data were generated such that the duration and anxiety covariates characterized three subgroups with differences in treatment effects. The cluster variable was used to introduce a random intercept that should be accounted for. The treatment outcome is an index of depressive symptomatology.


